---
title: "UMAP and SOM from Distance Matrices"
author: "C.E. Coombes, Zachary B. Abrams, Kevin R. Coombes"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{UMAP and SOM from Distance Matrices}
  %\VignetteKeywords{Mercator,self-organizing maps,uniofrm manifold approximationa nd projection}
  %\VignetteDepends{Mercator}
  %\VignettePackage{Mercator}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6, echo=TRUE)
```

# Introduction
We recently added visualizations based on Self-Organizing Maps (SOM) and
Uniform Manifold Approximation and Projection (UMAP) for objects in the
<tt>Mercator</tt> package. This addition relies on the core implementations
in the <tt>kohonen</tt> and <tt>umap</tt> packages, respectively. The main
challenge that we faced was that both of those implementations want to
receive the raw data as inputs, while <tt>Mercator</tt> objects only store
a distance matrix computed from the raw data. The purpose of this vignette
is, first, to describe how we worked around this restriction, and second,
to illustrate how to use these methods in the package.

# The Mercator Class
First we load the package.
```{r library}
suppressMessages( suppressWarnings( library(Mercator) ) )
```
Next, we load a "fake" set of synthetic continuous data that comes with the
Mercator package.
```{r fakedata}
set.seed(36766)
data(fakedata)
ls()
dim(fakedata)
dim(fakeclin)
```

# UMAP
Let's create a UMAP visualization directly from the raw data.
```{r udirect, fig.cap="UMAP visualization of fake data.", fig.width=7, fig.height=7}
library(umap)
udirect <- umap(t(fakedata))
plot(udirect$layout, main="Original UMAP")
```
We aren't going to do anything fancy with labels or colors in this plot; we just
want an idea of the main structure in the data. In particular, it looks like there
are seven clusters, divided into one group three and two groups of two each.

Next, we are going to create a <tt>Meractor</tt> object with a multi-dimensional
scaling (MDS) visualization. Inspired by the first plot, we will arbitrarily assign
seven groups. (This assignment uses the PAM algorithm internally.)
```{r mercury, fig.cap="MDS visualization from a distance matrix.", fig.width=7, fig.height=7}
mercury <- Mercator(dist(t(fakedata)), "euclid", "mds", 7)
plot(mercury, main = "MDS")
```
Interestingly, here we see a possibly different number of groups. But that's a
separate issue, and we don't want to distracted from our main thread.

We want to confirm that the actual raw data is not contained in <tt>mercury</tt>,
just the distance matrix.
```{r peek}
summary(mercury)
slotNames(mercury)
mercury@metric
mercury@palette
mercury@symbols
table(mercury@clusters)
class(mercury@view)
names(mercury@view)
class(mercury@view[["hclust"]])
class(D <- mercury@distance)
```

Next, we add a "umap" visualization to this object. 
```{r umapper, fig.cap="UMAP visualization from a distance matrix.", fig.width=7, fig.height=7}
mercury <- addVisualization(mercury, "umap")
plot(mercury, view = "umap", main="UMAP from distance matrix")
```
Notice that the <tt>plot</tt> method knows that we want to see the internal
"<tt>layout</tt>" component of the <tt>umap</tt> object. It also automatically
assigns axis names that (subliminally?) remind us that we computed this
visualization with UMAP. And it colors the points using the same values
from the initial PAM clustering assignments.

As an aside, we point out that the PAM clustering doesn't really match
the implicit UMAP clustering of the data.

# Euclideanization
We obviously constructed this plot just from the distance matrix, not from the
raw data. It neverheless has a structure essentially the same as the original one.
But how?

Well, internally, we use this function:
```{r me}
Mercator:::makeEuclidean
```
The first four lines of code inside this function are basically the algorithm
used by the <tt>cmdscale</tt> function to compute the eigenvalues need to create
its dimension reduction for visualization. The count of the number of nonzero
eigenvalues (which defines <tt>R</tt>) is careful to stay away from zero.  We do
this to avoid round-off errors. If we miss by one or two eignevalues, than the
final call to <tt>cmdscale</tt> will generate a confusing warning. That final line,
by the way, creates an embedding into Euclidean space that should preserve almost
all of the distances between pairs of points in the original data set.

To see that, let's carry out that step explicitly.
```{r newdist}
X <- Mercator:::makeEuclidean(D)
dim(X)
```
Now we compute distances between pairs of points in this new embedding.
```{r delta}
D2 <- dist(X)
delta <- as.matrix(D) - as.matrix(D2)
summary(as.vector(delta))
```
We see that the absolute errors between any pair of points in the two distance
computations are less than $10^{-12}$. And that is the fundamental reason that
we expect equivalent structures whether we apply UMAP to the original data or to
the re-embedded data arising from the distance matrix.

# Self-Organizing Maps
The same approach works to compute self-organizing maps from a distance matrix.
Here is the result using the data matrix.
```{r som, fig.cap="SOM visualizations from a distance matrix.", fig.width=7, fig.height=5}
mercury <- addVisualization(mercury, "som",
                            grid = kohonen::somgrid(6, 5, "hexagonal"))
plot(mercury, view = "som")
```
And here is the corresponding result using the original data.
```{r oldsom}
library(kohonen)
mysom <- som(t(fakedata), grid = somgrid(6, 5, "hexagonal"))
plot(mysom, type = "mapping")
```
This time, we will work a little harder to color the plot in the same way.
```{r colplot, fig.cap="SOM visualizations from a distance matrix.", fig.width=7, fig.height=5}
plot(mysom, type = "mapping", pchs=16, col=Mercator:::colv(mercury))
```

## Warning
Qualitatively, most of the SOM plots should be similar. The fundamental exception,
however, is the "codes" plot. This plot shows the patterns of intensities in
each of the underlying dimensions (averaged over the assigned objects in each node).
Because we have performed a multi-dimensional scaling analysis, we have changed the
underlying coordinate system. Here are the corresponding plots.
```{r codes, fig.cap="SOM codes.", fig.width=7, fig.height=5}
plot(mysom, type = "codes", main="Original", shape = "straight")
plot(mercury, view = "som", type="codes", main="After MDS")
```

# Conclusions
As we have seen, when we start with a Euclidean distance matrix, we can get
qualitatively equivalent results from the original data or from creating a
"realization" of the distance matrix in a large Euclidean space. The major
advantage, however, arises when we start with a completely different distance
metric. In that case, we cannot apply SOM or UMAP at all. But we can still embed
that distance matrix into a Eucldean space and then run SOM or UMAP.

# Appendix
This analaysis was performed in the following environment:
```{r si}
sessionInfo()
```